---
title: "Getting-And-Cleaning Course Proejct: CodeBook"
output: html_document
---

##Introduction
In this project, smartphone sensor signals (accelerometer and gyroscope) data 
from `http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones` 
is used as raw data. Note - data available under folder "Inertial Signals" is 
not used as test and training data produced from them is already available. 

The solution flow consists of three steps - Loading data, processing it, and 
writing tidy data generated to a file.

##Load raw data
In this section, we load the activity_labels, features, train, and test data from file system
```{r load-data, cache=TRUE}
activity_labels <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/activity_labels.txt")
names(activity_labels) <- c("activity_id", "activity_name")
#load features.txt
#561 x 2
features <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/features.txt")
names(features) <- c("feature_id", "feature_name")
#load train data
#7532 x 1
subject_train <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/train/subject_train.txt")
names(subject_train) <- c("subject_id")
#7532 x 561
X_train <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/train/X_train.txt")
#7532 x 1
y_train <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/train/y_train.txt")
names(y_train) <- c("activity")
#load test data
#2947 x 1
subject_test <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/test/subject_test.txt")
names(subject_test) <- c("subject_id")
#2947 x 561
X_test <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/test/X_test.txt")
#2947 x 1
y_test <- read.table("./getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/test/y_test.txt")
names(y_test) <- c("activity")
```

## Process the data to generate tidy data
In this section we perform following activities:
1. Merges the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement. 
3. Uses descriptive activity names to name the activities in the data set
4. Appropriately labels the data set with descriptive variable names. 
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
```{r process-data}
#Step1
#Combine X_train and y_train data to all_train
all_train <- cbind(subject_train, X_train, y_train)
#combine X_test and y_test to all_test
all_test <- cbind(subject_test, X_test, y_test)
#combine all_train and all_test data
#combine train and test data
all_data <- rbind(X_train, X_test)

#Step 2
#change columns names as feature names
names(all_data) <- as.vector(features[,2])
#select column names contaiming std and mean
all_data <- all_data[, as.vector(grep("std|mean", colnames(all_data), 
                                      value=FALSE))]
#Add subject and y (activity_id) data columns to above filtered data
subject_data <- rbind(subject_train, subject_test)
y_data <- rbind(y_train, y_test)
all_data <- cbind(subject_data, all_data, y_data)    

#Step 3
#Replace activity numbers with descrpitive activity names
library(plyr)
all_data$activity <- mapvalues(all_data$activity, c(1,2,3,4,5,6), 
                    as.character(activity_labels$activity_name))

#Step 4 - Make column names more descriptive
#Remove () after functions like std, mean
colnames(all_data) <- gsub("\\(\\)", "", colnames(all_data))
colnames(all_data) <- gsub("-X", "-X-axis", colnames(all_data))
colnames(all_data) <- gsub("-Y", "-Y-axis", colnames(all_data))
colnames(all_data) <- gsub("-Z", "-Z-axis", colnames(all_data))

#Step 5 - Create indendent tidy dataset with average of each variable for 
#each activity and each subject
library(dplyr, warn.conflicts=FALSE)
tidy_data <- group_by(all_data, subject_id, activity) 
tidy_data <- summarise_each(tidy_data, funs(mean))
```

##Store tidy data to a file
The cleaned and processed data is stored to file.
```{r store-data}
write.table(tidy_data, file="tidy_data_project.txt", row.names=FALSE, col.names=TRUE,
            sep="\t", quote=TRUE)
```

##Description of Variable names
- subject_id: identifies the subject who performed the activity
- activity: identifies name of the activity from list 
            WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING


In the rest of the virables -
XYZ-axis is used to denote 3-axial signals in the X, Y and Z directions.
mean and sd substrings in these variables denote average and standard deviation of variables.
These variables are measured in HZ units.

- tBodyAcc-XYZ
- tGravityAcc-XYZ
- tBodyAccJerk-XYZ
- tBodyGyro-XYZ
- tBodyGyroJerk-XYZ
- tBodyAccMag
- tGravityAccMag
- tBodyAccJerkMag
- tBodyGyroMag
- tBodyGyroJerkMag
- fBodyAcc-XYZ
- fBodyAccJerk-XYZ
- fBodyGyro-XYZ
- fBodyAccMag
- fBodyAccJerkMag
- fBodyGyroMag
- fBodyGyroJerkMag